---
title: "Proximal/Distal Axon Localization In Vitro"
output:
  html_document:
    df_print: paged
---
This script is designed to use the mean gray values from along an axon in an ICC image, where there is at least one channel in addition to DAPI. Bins the list of mean gray values and returns a proportion relative to the maximal value. 

TODO:
Also look at length of neurites at P1 vs P3
```{r}
library(dplyr)
library(ggplot2)
```

```{r}
project_dir <- "C:/Users/prv892/Documents/GitHub/amalgam"

plotprof_dir <- file.path(project_dir, "data/imaging/icc/plotprofile_quant")

# ideally change this to the illustrator_pdfs folder
plot_dir <-  file.path(project_dir, "data/imaging/icc/plotprofile_quant/plots")
```

```{r}

get_meta_from_files <- function(rbp_id) {
  rbp_meta <- data.frame(fname=file.path(plotprof_dir, rbp_id, list.files(file.path(plotprof_dir, rbp_id)))) %>%
    rowwise() %>%
    mutate(stage=strsplit(as.character(basename(fname)), "_")[[1]][2],
           neuron=strsplit(as.character(basename(fname)), "_")[[1]][3],
           channel=gsub(".csv", "", strsplit(as.character(basename(fname)), "_")[[1]][4]))

  return(rbp_meta)
}


```

```{r}
get_dapi_profiles <- function(rbp_meta) {
  
  dapi_files <- subset(rbp_meta, channel=="DAPI")$fname
  
  dapi_thresh <- rbp_meta$dapi_thresh[1]

  dapi_profiles <- data.frame(do.call(rbind, 
                     lapply(dapi_files, 
                            function(x) read.csv(x) %>% 
                              rowwise() %>% 
                              mutate(fname=basename(x)))))
  
  plot(ggplot(dapi_profiles, 
         aes(Gray_Value, color=fname)) + 
           geom_density() +
         theme(legend.position="None") +
         scale_x_log10() +
         geom_vline(xintercept = dapi_thresh))
}

```
```{r}
get_channel_meangray_bin <- function(file, dapi_file, dapi_thresh, num_bins=20) {

  
  meangray_file_colnames <- c("distance_microns", "gray_value", "is_axon")	

  # read in dapi channel
  res_dapi <- read.csv(dapi_file) 
  colnames(res_dapi) <- meangray_file_colnames

  # read in other channel
  res <- read.csv(file)
  colnames(res) <- meangray_file_colnames
  
  #print(ggplot(res_dapi, aes(distance_microns, gray_value)) + geom_point() + geom_hline(yintercept=400))
  
  # perform the axon computation
  axon_mask_dapi <- res_dapi %>% 
    subset(is_axon == 1 & gray_value < dapi_thresh) %>% 
    dplyr::select(distance_microns)

  
  res_axon <- res %>% inner_join(axon_mask_dapi)
  #print(ggplot(res, aes(distance_microns, gray_value)) + geom_point())



  axon_length_bins <- seq(0, max(res_axon$distance_microns), 
                          by=max(res_axon$distance_microns)/num_bins)
  axon_length_bins_offset <- axon_length_bins + (max(res_axon$distance_microns)/num_bins)/2
  
  
  res_axon$bin <- .bincode(res_axon$distance_microns, 
                           axon_length_bins, include.lowest = TRUE)
  
  res_axon_offset <- res_axon
  res_axon_offset$bin <- .bincode(res_axon$distance_microns,
                                  axon_length_bins_offset, include.lowest = TRUE)+0.5
  
  res_axon <- data.frame(rbind(res_axon, res_axon_offset))
  
  res_axon_binvals <- res_axon %>% 
    group_by(bin) %>% 
    summarize(bin_value_mean=mean(gray_value),
              bin_stdev=sd(gray_value))
  
  res_axon_binvals$bin_name <- sprintf("Bin %s", res_axon_binvals$bin)
  res_axon_binvals$bin_name <- factor(res_axon_binvals$bin_name, 
                                      levels=sprintf("Bin %s", 1:num_bins))
  
  # perform the soma computation
  soma_mask_dapi <- res_dapi %>%
    subset(is_axon == 0) %>% 
    dplyr::select(distance_microns)
  
  res_soma <- res %>% inner_join(soma_mask_dapi)
  res_soma_meangray <- data.frame(bin=0, bin_name="Soma",
                                  bin_value_mean=mean(res_soma$gray_value),
                                  bin_stdev=sd(res_soma$gray_value))


  res_binvals <- data.frame(rbind(res_axon_binvals, res_soma_meangray))
  
  
  res_binvals$fname <- file

  return(res_binvals)
  
}

get_neuron_channel_meangray_bin <- function(neuron_meta) {
  dapi_file <- subset(neuron_meta, channel == "DAPI")$fname
  dapi_thresh <- neuron_meta$dapi_thresh[1]
  
  channel_bins <- data.frame(do.call(rbind, 
                          lapply(neuron_meta$fname, 
                                 function(x) get_channel_meangray_bin(x, dapi_file, dapi_thresh))))
  

  return(channel_bins)

}

calculate_all_neurons <- function(rbp_meta) {
  rbp_bin_result <- data.frame(do.call(rbind,
                            tibble(rbp_meta) %>% 
    group_by(stage, neuron) %>%
    group_map(~get_neuron_channel_meangray_bin(.x)))) %>%
    inner_join(rbp_meta, by="fname")
  
  return(rbp_bin_result)
}

calculate_stats_all_neurons <- function(rbp_neuron_bins) {
  bin_stats <- rbp_neuron_bins %>% 
    subset(!is.na(bin_stdev)) %>%
    group_by(stage, channel, bin, bin_name) %>%
    summarize(mean_gray_value=mean(bin_value_mean),
              stderr_gray_value=(sqrt(sum(bin_stdev^2))/length(bin_stdev))/sqrt(length(bin_stdev)),
              num_values_bin=length(bin_value_mean))
  
  return(bin_stats)
    
}

plot_soma <- function(neuron_bins, plot_savefile, h=80, w=80) {
  
  for (chan in c("FITC", "TRITC", "Cy5")) {
    plt_mean <- ggplot(subset(neuron_bins, bin_name == "Soma" & channel==chan ),
       aes(stage, bin_value_mean, fill=stage)) + geom_boxplot() +
                 ggtitle(sprintf("Channel: %s, Stat: mean", chan)) +
      theme(legend.position="None") +
      theme_classic()

    
    plot(plt_mean)
    
    ggsave(gsub(".pdf", sprintf("_%s.pdf", chan), plot_savefile),
           plot=plt_mean,
           height=h,
           width=w,
           units="mm",
           useDingbats=FALSE)


  }
  
}


plot_axon <- function(neuron_bin_stats, plot_savefile, h=160, w=80) {
  
  for (chan in c("FITC", "TRITC", "Cy5")) {
    
    df_to_plot <- subset(neuron_bin_stats, bin_name != "Soma" & channel == chan & num_values_bin >= 5)
    # need to reorder the bins to get proximal to distal in right order
    maxval_bins <- max(df_to_plot$bin)
    df_to_plot$bin <- maxval_bins+1-df_to_plot$bin
    
    plt <- ggplot(df_to_plot,
                  aes(bin, mean_gray_value, group=stage, color=stage)) +
      geom_line() +
      geom_ribbon(data=df_to_plot,
                  mapping=aes(x=bin, 
                              ymin=mean_gray_value-stderr_gray_value,
                              ymax=mean_gray_value+stderr_gray_value,
                              fill=stage,
                              group=stage),
                  alpha=0.5) +
      ggtitle(sprintf("Channel: %s, Stat: mean", chan)) +
      theme_classic() +
      coord_flip()
    
    plot(plt)
    
    ggsave(gsub(".pdf", sprintf("_%s.pdf", chan), plot_savefile), plot=plt,
           device="pdf",
           height=h, 
           width=w,
           units="mm",
           useDingbats=FALSE)


  }
  
}
```

Calculate the axon stats and plot
Weird bump in the middle of P1 Cpeb4 could be real? not a dapi signal. Look at neuron 5 image, for example.
```{r}
cpeb4_meta <- get_meta_from_files("Cpeb4")
cpeb4_meta$dapi_thresh <- 500
get_dapi_profiles(cpeb4_meta)


cpeb4_neuron_bins <- suppressMessages(calculate_all_neurons(cpeb4_meta))

plot_soma(cpeb4_neuron_bins, file.path(plot_dir, "Cpeb4_soma_intensities.pdf"))

cpeb4_neuron_bin_stats <- calculate_stats_all_neurons(cpeb4_neuron_bins)
plot_axon(cpeb4_neuron_bin_stats, file.path(plot_dir, "Cpeb4_axon_profiles.pdf")) 

```
```{r}
rbms1_meta <- get_meta_from_files("Rbms1")
rbms1_meta$dapi_thresh <- 500
get_dapi_profiles(rbms1_meta)

rbms1_neuron_bins <- suppressMessages(calculate_all_neurons(rbms1_meta))

plot_soma(rbms1_neuron_bins, file.path(plot_dir, "Rbms1_soma_intensities.pdf"))

rbms1_neuron_bin_stats <- calculate_stats_all_neurons(rbms1_neuron_bins)
plot_axon(rbms1_neuron_bin_stats, file.path(plot_dir, "Rbms1_axon_profiles.pdf")) 

```


```{r}
pum1_meta <- get_meta_from_files("Pum1")
pum1_meta$dapi_thresh <- 500
get_dapi_profiles(pum1_meta)

pum1_neuron_bins <- suppressMessages(calculate_all_neurons(pum1_meta))

plot_soma(pum1_neuron_bins, file.path(plot_dir, "Pum1_soma_intensities.pdf"))

pum1_neuron_bin_stats <- calculate_stats_all_neurons(pum1_neuron_bins)
plot_axon(pum1_neuron_bin_stats, file.path(plot_dir, "Pum1_axon_profiles.pdf")) 

```

```{r}
pum2_meta <- get_meta_from_files("Pum2")
pum2_meta$dapi_thresh <- 500
get_dapi_profiles(pum2_meta)

pum2_neuron_bins <- suppressMessages(calculate_all_neurons(pum2_meta))

plot_soma(pum2_neuron_bins, file.path(plot_dir, "Pum2_soma_intensities.pdf"))

pum2_neuron_bin_stats <- calculate_stats_all_neurons(pum2_neuron_bins)
plot_axon(pum2_neuron_bin_stats, file.path(plot_dir, "Pum2_axon_profiles.pdf")) 

```

```{r}
wave1_meta <- get_meta_from_files("WAVE1")
wave1_meta$dapi_thresh <- 500
get_dapi_profiles(wave1_meta)

wave1_neuron_bins <- suppressMessages(calculate_all_neurons(wave1_meta))

wave1_neuron_bins <- wave1_neuron_bins %>%
  rowwise() %>%
  mutate(neuron=sprintf("%s_%s", stage, neuron),
         stage="P3")
wave1_neuron_bin_stats <- calculate_stats_all_neurons(wave1_neuron_bins)
plot_axon(wave1_neuron_bin_stats, file.path(plot_dir, "Wave1_axon_profiles.pdf")) 

```
